---
title: "Relandscaping Cleaning Template"
format: html
editor: visual
---

## Relandscaping Data Cleaning: Stomatal Conductance

Winter 2024 - MCRS Advanced Quantitative Methods Class Project

Team #: 2

Team Members: Luke, Darien, Izzy, Yvonne

### Load Packages

```{r}
#install.packages(tidyverse)
library(tidyverse)
library(ggplot2)
library(readr)
library(readxl)
```

### Import Data Files

```{r}
StomCon_Winter <- read_excel("StomCon_Winter 2019 GH Relandscaping Stomatal Conductance Data.xlsx", col_types = "text")
            # col_names = FALSE, 
             # col_names = TRUE,
             #col_types = TRUE, 
            #col_types = "text" 

```

```{r}
StomCon_Fall <- read_excel("StomCon_Fall 2018 GH Relandscaping Stomatal Conductance Data.xlsx", col_types = "text")
```

### Create Primary

Reformat the data from each of your imported spreadsheets to have common columns and column formats so you can bind them into one object.

#### Data Frame #1: Fall 2018 data

```{r}
#No changes to column names
```

#### Data Frame #2: Winter 2019

```{r}
StomCon_Winter_Clean <- StomCon_Winter |>
  select(1:10)
```

#### Complete Raw Data

Bind your cleaned data frames together into one object. Clean up any remaining column names or formats.

```{r}
StomCon_Combined <- bind_rows(StomCon_Fall, StomCon_Winter_Clean)
```

### Cleaning

Go through your raw data frame column by column to find and correct issues.

#### Issue #1: Get rid of the first row

```{r}
StomCon_Combined_1 <- StomCon_Combined|>
  filter(Block != '43103.0')
```

#### Issue #2: Select columns

```{r}
StomCon_Combined_2 <- StomCon_Combined_1|>
  select(-Observers, -Notes)
```

etc.

#### Issue #3: Change Non-Native to NonNative

```{r}
StomCon_Combined_3 <- StomCon_Combined_2|>
  mutate(Treatment = case_when(
   Treatment == "Non-Native" ~ "NonNative",
   is.character(Treatment) == TRUE ~ Treatment
 ))
View(StomCon_Combined)
```

#### Issue #4: Change Date and Get rid of NA data

```{r}
StomCon_Combined_4 <- StomCon_Combined_3 |>
  mutate(Date = as.Date(as.numeric(Date), origin = "1900-01-01")) |>
  na.omit() |>
  filter(`Meas. 1` != "-", `Meas. 2` != "-", `Meas. 3` != "-")

#View(StomCon_Combined_4)
```

#### Issue #5: Column names

```{r}
StomCon_Clean <- StomCon_Combined_4 |>
  rename(Species = "Plant Species" ) |>
  rename(Tag_Num = "Tag Number")|>
  rename(StomCon1 = "Meas. 1" , StomCon2 = "Meas. 2" , StomCon3 = "Meas. 3")

  
```

### Save your cleaned and complete Primary data frame to .csv and .Rdata and save the file as VARIABLE_Primary.csv or similar.

```{r}

```
